---
title: "Assessment_2"
author: "C Rusling"
date: "2024-11-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Aim

The aim of this report is to explore the impact of the COVID-19 pandemic on the percentage of children reported to have at least one developmental concern at 27-30 months.

## Load Packages

\`

```{r load packages}
library(tidyverse)
library(viridis)   
library(plotly) 
library(here)
library(ggplot2)
library(ggbump)
library(hrbrthemes)
library(sf)

```

## Data Acquisition

Data file on developmental concerns at 27 to 30 months from ScoPho including all available years for Scotland, health boards, and intermediate data zones.

```{r read in data}
dev_data <- read_csv(here("./inputs/dev_concerns_all.csv"))
shapefile <- read_sf(here("./Inputs/Shape file-SG_NHS_HealthBoards_2019"))
```

## Data Cleaning

```{r glimpse data for Developmental Concerns at 27-30 months}
dev_data %>% glimpse
```

Filter data to health boards and keep only the variables of interest, including the area name, area code, year, and measure.

```{r filter variables}

hb_dev_data <- dev_data %>%  
  filter(area_type == "Health board") %>%
  select(area_name, area_code, year, measure) 
glimpse(hb_dev_data)

```

Rename variables for ease of plotting.

```{r rename variables}
hb_dev_data <- hb_dev_data %>%
  mutate(area_name = gsub("NHS", "", area_name)) %>%
  rename(health_board = area_name, percent_concern = measure) 
```

## Data Analysis

Find health boards with the biggest difference in percentage of concern pre-COVID (2019) and post-COVID (2021) to highlight areas of concern in plots.

```{r find difference}

#Explore the data 

hb_dev_data %>%
  group_by(health_board) %>%
  summary(percent_concern)

#Calculate the difference of concern year over year in each health board 

diff_data_all <- hb_dev_data %>%
  group_by(health_board) %>%
  arrange(year) %>%
  mutate(difference = percent_concern - lag(percent_concern, default = first(percent_concern))) %>%
  mutate(cumm_diff = cumsum(difference)) %>%
  arrange(health_board)

# Isolate years of interest 

diff_data <- hb_dev_data %>%
  group_by(health_board) %>%
  filter(year == 2019 | year== 2020 | year == 2021) %>%
  arrange(year) %>%
  mutate(difference = percent_concern - lag(percent_concern, default = first(percent_concern))) %>%
  mutate(cumm_diff = cumsum(difference)) %>%
  arrange(health_board) 
```

Rank the percentage of concern by health board for 2019, 2020, 2021 and rank the difference of percentage from 2019 to 2021 by health board. This will help illustrate areas of concern that may be over looked if based on percentage of concern alone. 

```{r rankings}
# See rankings
diff_rank <- diff_data %>%
  group_by(year) %>%
  mutate((rank_diff = rank(cumm_diff)), (rank_per = rank(percent_concern))) %>%
  rename(rank_diff = "(rank_diff = rank(cumm_diff))", rank_per = "(rank_per = rank(percent_concern))")

# Filter for 2021
diff_rank_2021 <- diff_rank %>%
  filter(year== 2021) %>%
  select(1,7,8) 

tibble(diff_rank_2021)
```

```{r geospatial data}
# Merge geospatial and numeric data for mapping
dev_2019 <- hb_dev_data %>%
  filter(year == 2019)
sf_dev_2019 <- shapefile %>%
  left_join(dev_2019, by = c("HBCode" = "area_code"))
dev_2020 <- hb_dev_data %>%
  filter(year == 2020)
sf_dev_2020 <- shapefile %>%
  left_join(dev_2020, by = c("HBCode" = "area_code"))
dev_2021 <- hb_dev_data %>%
  filter(year == 2021)
sf_dev_2021 <- shapefile %>%
  left_join(dev_2021, by = c("HBCode" = "area_code"))
```


## Communication

```{r line plot}

hb_dev_data %>%
 ggplot( aes(x=year, y=percent_concern, group=health_board, fill=health_board)) +
    geom_area(show.legend = FALSE) +
    scale_color_viridis(discrete = TRUE) +
    theme_classic() +
  labs(title = "Developmental Concerns at 27-30 Months", 
       subtitle = "Percentage of Concern by Health Board",
       y = "percent",
       x = " ") +
  facet_wrap(~health_board) 
 
```

```{r bar plot}
diff_data %>%
  filter(year %in% c(2021)) %>%
  mutate(year=as.factor(year)) %>%
  mutate(health_board=as.factor(health_board)) %>%
ggplot(aes(fill= reorder(health_board, +`cumsum(difference)`), y=`cumsum(difference)`, x= reorder(health_board, +`cumsum(difference)`))) + 
  geom_col() + scale_color_viridis(discrete = TRUE) +
  theme_classic() +
  labs(title = "Developmental Concerns at 27-30 Months", 
       subtitle = "Difference in Percentage of Concern 2019-2021",
       x = "", 
       y = "percentage of change",
       fill = "Health Board") +
  theme(axis.ticks.x = element_blank(),
       axis.text.x = element_blank() )
       


```
```{r}
diff_rank %>%
  filter(year == 2021) %>%
  ggplot(aes(x = percent_concern, y = cumm_diff, size = rank_diff, colour = health_board)) +
           geom_point(alpha = 0.7) +
  scale_size(range = c(1, 14), guide = FALSE) +
  scale_colour_viridis(discrete = TRUE) +
  theme_classic()
  
```



```{r}
map_2021 <- sf_dev_2021 %>%
  ggplot(aes(fill = percent_concern)) + 
  geom_sf(colour = NA) + # Adding 'colour = NA' removes boundaries 
  # around each LA
  scale_fill_viridis("percent_concern") + # Changes our colour scale to 
  # viridis and prettifies our fill title in the legend
  theme_void() # removes unnecessary background grid and axes
map_2021
```
## Summary
Who is your target audience, and what is the key message you are aiming to get across (with your chosen visualisations)?
What data did you use to produce your visualisation? What are the data limitations?
What are the strengths and limitations of the approach you took for your visualisation?

## References